---
format:
  typst:
    papersize: a0
    margin:
      x: 5cm
      y: 5cm
    mainfont: "Arial"
    fontsize: 20pt
---

```{=typst}
#set text(font: "Arial", size: 20pt)

#show heading.where(level: 1): it => block(
  text(font: "Arial", size: 96pt, weight: "bold", it.body),
  below: 12pt
)

#show heading.where(level: 2): it => block(
  text(font: "Arial", size: 44pt, weight: "regular", it.body),
  below: 24pt
)

#show heading.where(level: 3): it => block(
  text(font: "Arial", size: 40pt, weight: "bold", it.body),
  below: 40pt
)

#show figure.caption: it => text(font: "Arial", size: 24pt, it)
```

# _tximeta_ support for mixed reference sets

## Michael I. Love^1^, Charlotte Soneson^2^

^1^Department of Genetics, UNC-Chapel Hill, Chapel Hill, NC, US \
^2^Computational Biology Platform, Friedrich Miescher Institute for Biomedical Research, Basel, Switzerland

---

```{=typst}
#columns(2, gutter: 1.5cm)[
```

## Introduction

RNA-seq quantification tools such as _salmon_ and _oarfish_ operate against
an index built from reference transcript sequences. A key feature of
these tools is that they record a digest of those sequences at
index-build time. _tximeta_ reads that digest at import time to automatically
identify the provenance of the reference. It then attaches the appropriate
transcript and genome metadata to the resulting _SummarizedExperiment_,
in the `rowData` and `rowRanges`.
This workflow eliminates the need for users to manually track which annotation
release was used, and to manually reconstruct the matching _TxDb_.
This model for attaching data provenance 
works regardless of the specific steps or actions taken 
by the initial analyst performing quantification; 
no extra steps are required.

The long-read quantifier _oarfish_ and _tximeta_ recently 
extend this model to
*mixed reference* indexing: a single index may combine a set of
*annotated* transcripts (e.g. from GENCODE or Ensembl) with a set of
*novel* transcripts (e.g. de novo assembled or fusion sequences not
present in the reference annotation). _tximeta_ v1.28 introduced the 
following functions:

  - `importData()`
  - `inspectDigests()`
  - `updateMetadata()`
  - `linkedTxpData()`

These allow the user to check each sub-index digest independently ,
add linked metadata, and populate transcript metadata wherever 
a match is found.
`linkedTxpData()` allows simple tabular metadata on transcripts 
to be stored in the file cache, where previously only GTF-based 
metadata was supported in _tximeta_ via `linkedTxome()`.

## Setup

Three replicates from the SGNEx H9 cell line, quantified with oarfish
against a mixed reference: GENCODE v48 annotated transcripts plus a set
of novel transcripts.

```{r}
#| include: false
library(BiocFileCache)
bfc <- BiocFileCache()
# remove the linked txpData
res <- bfcquery(bfc, "linkedTxpData")
if (nrow(res) > 0) {
  bfcremove(bfc, rids = res$rid)
}
# remove the txpData for 43158f
res <- bfcquery(bfc, "txpdata-43158f")
if (nrow(res) > 0) {
  bfcremove(bfc, rids = res$rid)
}
```

```{r}
#| message: false
library(tximeta)
library(SummarizedExperiment)

dir <- system.file("extdata/oarfish", package = "tximportData")
names <- paste0("rep", 2:4)
files <- file.path(dir, paste0("sgnex_h9_", names, ".quant.gz"))
coldata <- data.frame(files, names)
```

## Import

`importData()` reads the counts and stores the index digests in metadata,
returning an un-ranged `SummarizedExperiment`.

```{r}
se <- importData(coldata, type = "oarfish")
```

## Inspect digests

`inspectDigests()` checks each index digest against tximeta's registries.
The annotated index matches GENCODE v48 in the pre-computed hash table;
the novel index has no match yet.

```{r}
inspectDigests(se)
```

## Update metadata

`updateMetadata()` populates `rowData` for any transcripts whose index
has a known match. Here the annotated GENCODE v48 transcripts receive
`tx_name`, `gene_id`, and other columns; the novel transcripts are left
with `NA` metadata and `index = "novel"`.

```{r}
#| collapse: true
se_updated <- updateMetadata(se)
```

```{r}
n <- nrow(se_updated)
rowData(se_updated)[ c(1:2, (n-1):n), ]
```

To also attach genomic ranges:

```{r}
#| eval: false
se_ranges <- updateMetadata(se, ranges = TRUE)
rowRanges(se_ranges)
```

## Supplying novel transcript metadata

Novel transcript metadata can be provided directly via the argument `txpData`, 
or one can use `linkedTxpData()` to register them persistently, using the file cache. 
`txpData` can take tabular data or _GRanges_ as input, 
here we show with ranges:

```{r}
#| include: false
novel <- data.frame(
  seqnames = paste0("chr", rep(1:22, each = 500)),
  start    = 1e6 + 1 + 0:499 * 1000,
  end      = 1e6 + 1 + 0:499 * 1000 + 1000 - 1,
  strand   = "+",
  tx_name  = paste0("novel", 1:(22 * 500)),
  gene_id  = paste0("novel_gene", rep(1:(22 * 10), each = 50))
)
```

```{r}
#| collapse: true
library(GenomicRanges)
head(novel, 3)
novel_gr <- as(novel, "GRanges")
names(novel_gr) <- novel$tx_name
```

```{r}
#| collapse: true
se_full <- updateMetadata(se, txpData = novel_gr, ranges = TRUE)
n <- nrow(se_full)
rowData(se_full)[ c(1:2, (n-1):n), ]
table(mcols(se_full)$index)
```

## Persistent metadata via linking

Rather than supplying `txpData` each time, novel transcript metadata can be
registered once in the file cache with `makeLinkedTxpData()`. The novel index
digest is obtained from the `inspectDigests()` output or from the 
oarfish output directly.

```{r}
#| collapse: true
makeLinkedTxpData(
  digest = "43158f2c8e88e3acd77c22aee557625a6f1b6a5038cfc7deb5e64903892d8070",
  digestType = "sha256",
  indexName = "my_novel_txps",
  txpData = novel_gr,
  source = "novel",
  organism = "Homo sapiens",
  release = "v1",
  genome = "GRCh38"
)
```

After registration, `inspectDigests()` reflects the match, and
`updateMetadata()` finds and applies the linked metadata.

```{r}
inspectDigests(se)
```

```{r}
#| collapse: true
se_linked <- updateMetadata(se)
```

```{r}
n <- nrow(se_linked)
rowData(se_linked)[ c(1:2, (n-1):n), ]
```

## Next steps

**Backporting to _salmon_.** The mixed-reference digest workflow currently
targets _oarfish_, which natively records separate digests for each
sub-index in its metadata JSON. _salmon_ does not yet emit per-sub-index
digests, so the `importData()` / `inspectDigests()` / `updateMetadata()`
pipeline is not yet available to _salmon_ users who combine annotated and
novel sequences in a single index. Extending support will require either
a change to _salmon_'s metadata output, or a preprocessing step that
records the sub-index digests separately before indexing.

**Quantification over subsets of the annotated transcript set.** A
common use case is to quantify against a subset of the annotated
reference, e.g. transcripts with evidence of expression in a particular
tissue. Digest-based provenance breaks down here because
the subset has a different hash from any entry in the pre-computed table,
so _tximeta_ cannot identify it automatically.

One solution is a high-level framework that provides uniform,
simple interfaces across _salmon_ and _oarfish_. Such a framework could
manage subsetting of the annotated reference prior to indexing: it would
compute and store digests of the original and subset FASTA in a standard
location, then propagate these digests into the tool's index metadata and
quantification output files during the indexing and quantification steps.

```{=typst}
]
```
