---
format:
  typst:
    papersize: a0
    margin:
      x: 10cm
      y: 10cm
    mainfont: "Arial"
    fontsize: 24pt
---

```{=typst}
#set text(font: "Arial", size: 24pt)

#show heading.where(level: 1): it => block(
  text(font: "Arial", size: 96pt, weight: "bold", it.body),
  below: 12pt
)

#show heading.where(level: 2): it => block(
  text(font: "Arial", size: 44pt, weight: "regular", it.body),
  below: 24pt
)

#show heading.where(level: 3): it => block(
  text(font: "Arial", size: 40pt, weight: "bold", it.body),
  below: 40pt
)

#show figure.caption: it => text(font: "Arial", size: 24pt, it)
```

# _tximeta_ support for mixed reference sets

## Michael I. Love^1^, Charlotte Soneson^2^

^1^Department of Genetics, UNC-Chapel Hill, Chapel Hill, NC, US \
^2^Computational Biology Platform, Friedrich Miescher Institute for Biomedical Research, Basel, Switzerland

---

```{=typst}
#columns(2, gutter: 1.5cm)[
```

## Introduction

RNA-seq quantification tools such as salmon and oarfish operate against
an index built from reference transcript sequences. A key feature of
tximeta is that it records a SHA-256 digest of those sequences at
index-build time, and uses that digest at import time to automatically
identify the provenance of the reference and attach the appropriate
transcript and genome metadata to the resulting `SummarizedExperiment`.
This eliminates the need for users to manually track which annotation
release was used and reconstruct the matching `TxDb` or GTF.

Long-read quantifiers such as oarfish extend this model to
*mixed-reference* indexing: a single index may combine a set of
*annotated* transcripts (e.g. from GENCODE or Ensembl) with a set of
*novel* transcripts (e.g. de novo assembled or fusion sequences not
present in the reference annotation). tximeta v1.28 introduced
`importData()`, `inspectDigests()`, and `updateMetadata()` to handle
this case, checking each sub-index digest independently and populating
transcript metadata wherever a match is found.

## Setup

Three replicates from the SGNEx H9 cell line, quantified with oarfish
against a mixed reference: GENCODE v48 annotated transcripts plus a set
of novel transcripts.

```{r}
#| message: false
library(tximeta)
library(SummarizedExperiment)

dir <- system.file("extdata/oarfish", package = "tximportData")
names <- paste0("rep", 2:4)
files <- file.path(dir, paste0("sgnex_h9_", names, ".quant.gz"))
coldata <- data.frame(files, names)
```

## Import

`importData()` reads the counts and stores the index digests in metadata,
returning an un-ranged `SummarizedExperiment`.

```{r}
se <- importData(coldata, type = "oarfish")
se
```

## Inspect digests

`inspectDigests()` checks each index digest against tximeta's registries.
The annotated index matches GENCODE v48 in the pre-computed hash table;
the novel index has no match yet.

```{r}
inspectDigests(se)
```

## Update metadata

`updateMetadata()` populates `rowData` for any transcripts whose index
has a known match. Here the annotated GENCODE v48 transcripts receive
`tx_name`, `gene_id`, and other columns; the novel transcripts are left
with `NA` metadata and `index = "novel"`.

```{r}
#| eval: false
se_updated <- updateMetadata(se)
mcols(se_updated)
```

To also attach genomic ranges (requires downloading the GENCODE v48 GTF):

```{r}
#| eval: false
se_ranges <- updateMetadata(se, ranges = TRUE)
se_ranges
```

## Supplying novel transcript metadata

Novel transcript ranges can be provided directly via `txpData`, without
needing to register them persistently.

```{r}
library(GenomicRanges)

novel <- data.frame(
  seqnames = paste0("chr", rep(1:22, each = 500)),
  start    = 1e6 + 1 + 0:499 * 1000,
  end      = 1e6 + 1 + 0:499 * 1000 + 1000 - 1,
  strand   = "+",
  tx_name  = paste0("novel", 1:(22 * 500)),
  gene_id  = paste0("novel_gene", rep(1:(22 * 10), each = 50)),
  type     = "protein_coding"
)
novel_gr <- as(novel, "GRanges")
names(novel_gr) <- novel$tx_name
```

```{r}
#| eval: false
se_full <- updateMetadata(se, txpData = novel_gr, ranges = TRUE)
mcols(se_full)
table(mcols(se_full)$index)
```

## Next steps

**Backporting to salmon.** The mixed-reference digest workflow currently
targets oarfish, which natively records separate digests for each
sub-index in its metadata JSON. Salmon does not yet emit per-sub-index
digests, so the `importData()` / `inspectDigests()` / `updateMetadata()`
pipeline is not yet available to salmon users who combine annotated and
novel sequences in a single index. Extending support will require either
a change to salmon's metadata output, or a preprocessing step that
records the sub-index digests separately before indexing.

**Quantification over subsets of the annotated transcript set.** A
common use case is to quantify against a subset of the annotated
reference — for example, transcripts on the primary assembly only, or a
biotype-filtered set. Digest-based provenance breaks down here because
the subset has a different hash from any entry in the pre-computed table,
so tximeta cannot identify it automatically.

A natural solution is a high-level framework that provides uniform,
simple interfaces across salmon and oarfish. Such a framework could
manage subsetting of the annotated reference prior to indexing: it would
compute and store the SHA-256 digest of the subset FASTA in a standard
location, then propagate that digest into the tool's index metadata and
quantification output files during the indexing and quantification steps.
tximeta would then be able to look up that digest in a locally maintained
registry, recovering full provenance without any manual linking step by
the user.

### References

::: {#refs}
:::

```{=typst}
]
```
